---
title: "Week 2 - Getting and Cleaning Data"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

## Parsing SQL data

- All the goodness of SQL
- Strucutre: Databases, tables, fields
- Rows as 'Records'

### Getting a SQL genome from UCSC

```{r}
## Make the obj into a handle
ucscDb <- dbConnect(MySQL(), 
                    user="genome",
                    host="genome-mysql.cse.ucsc.edu")

## Query the database
## "show databases" - mySQL command
result <- dbGetQuery(ucscDb, "show databases;")
dbDisconnect(ucscDb)
```

Show all databases from UCSC:

```{r}
result
```

### Connecting to hg19

```{r}
### Acessing the hg db
hg19 <- dbConnect(MySQL(),
                  user = "genome",
                  db = "hg19",
                  host = "genome-mysql.cse.ucsc.edu")

allTables <- dbListTables(hg19)
length(allTables)
```
```{r}
## Components of human genome!
allTables[1:5]
```

### Getting dimensions of a specific table

```{r}
dbListFields(hg19, "affyU133Plus2")
```

- Querying with SQL commands:

```{r}
## How many spots in the Affymetrix U133+2? (rows)
dbGetQuery(hg19, "select count(*) from affyU133Plus2")
```

#### Getting an actual SQL table

```{r}
affyData <- dbReadTable(hg19, "affyU133Plus2")
head(affyData)
```

- Getting a specific subset

```{r}
query <- dbSendQuery(hg19,
                     "select * from affyU133Plus2
                     where misMatches between 1 and 3")

## Fetching the query
affyMis <- fetch(query)
quantile(affyMis$misMatches)
```

```{r}
## Querying the first 10
affyMisSmall <- fetch(query, n = 10)
dbClearResult(query)
```

- A small table:

```{r}
dim(affyMisSmall)
```

- The entire table:

```{r}
dim(affyMis)
```

- Remember to close the connection:

```{r}
dbDisconnect(hg19)
```

# Reading from HDF5

- Store large sets
- Supports a range of types
- Hierarchical
- Groups: zero or more sets and metadata
  - Group headers: name and list of attributes
  - Group symbol table: list of objects in a group
- Datasets: N dim array of elements with metadata
  - Header with name, datatype, dataspace, storage layout
  - Data array with data

## Installing R-HDF5

```{r}
# If you dont already have:
# install.packages("BiocManager")
# BiocManager::install("rhdf5")
library(rhdf5)
```

If it worked:

```{r}
created = h5createFile("example.h5")
created
```

Creating groups:

```{r}
created <- h5createGroup("example.h5", "foo")
created <- h5createGroup("example.h5", "baa")
created <- h5createGroup("example.h5", "foo/foobaa")
h5ls("example.h5")
```

Writing to groups:

Obs - may fit matrices, multidim arrays, quite anything.

```{r}
A = matrix(1:10, nr=5, nc=2)
h5write(A, "example.h5", "foo/A")

B = array(
  seq(0.1, 2.0, by=0.1),
  dim=c(5,2,2)
)
h5write(B, "example.h5", "foo/foobaa/B")

h5ls("example.h5")
```

Writing a dataset:

```{r}
df = data.frame(1L:5L,
                seq(0, 1, length.out=5),
                c("ab", "cde", "fghi", "a", "s"),
                stringsAsFactors = F)


```








