---
title: "Week 1 - R Nuts and Bolts"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	cache = TRUE,
	comment = NA,
	include = FALSE,
	prompt = FALSE,
	include = TRUE,
	tidy = TRUE
)
opts_knit$set(width=75)
```


## Preamble

### Additional packages and notes

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

## History of R

- First written in Fortran, then in C
- Environment for basic analysis, and then programming

## Features i didn't know of

- Identical to S
- Run on anything (even Playstation 3)
- Constant releases
- User turns into a programmer

## Free software?

1. Run the program
2. Study the program and source code
3. Freedom to redistribute
4. Freedom to improve the program

## Drawbacks

1. 40-year old technology
2. 3D/dynamic graphics sucks
3. Functionality is based on demand
4. Objects are (usually) stored in physical memory
5. Slow if you don't know C++can 

## Asking questions

Pretty basic, i'm a master forumer already.

## Code: Input and Evaluation

```{r include=TRUE}
## Assignment operators
x = 1
y <- 2
print(x)
print(y)

## Incomplete expression
# ex:
# x <-
```

Evaluation: whenever some object is called, it is auto-printed:

```{r}
print(x)
```

Let's try a sequence:

```{r}
seq = 1:20
print(seq)
```


## R Objects and Attributes

- Everything is an object
- Atomic types:
  - Char
  - Num
  - Int
  - Complex
  - Logicals
- The most basic object is a vector
- Lists are represented as vector, but may contain anything inside

### Numbers

- Num objects
- An integer must be specified with L, such as: `1L`
- Other numbers:
  - `Inf` is infinity
  - `NaN` is "not a number", or missing value
  
### Attributes

R objects can have attributes:

- names, dimnames
- dimensions
- class
- length
- anything user-defined
- check with `attributes()`

### Data types: vectors and functions

#### c() - concatenates values

Creates vectors with the desired content

```{r}
x = c(0.5, 0.6)
x = c(2L, 4L)
x = 9:29
x = c(1+0i, 2+4i)
```

#### vector() - creates vectors

```{r}
x = vector("numeric", length = 15)
print(x)
```

#### Mixing objects and coercion

when mixing objects in a vector, the least enforced type is coerced, from boolean, to integer, to character

#### Explicit coercion: as.type()

We can force a class to be whatever type we want explicitly:

```{r}
x = 0:5
as.numeric(x)
as.logical(x)
as.character(x)
as.complex(x)
```

Nonsensical coercion results in NAs:

```{r}
x = c("a","b","c")
as.numeric(x)
as.logical(x)
as.complex(x)
```

### Lists

Super vectors that can carry anything as an indexed element

```{r}
x = list(1, "a", T, 1+4i)
print(x)
```

### Matrices

Vectors with a dimension attribute, an integer

```{r}
m = matrix(nrow = 2, ncol = 3)
m
```
```{r}
dim(m)
```
```{r}
attributes(m)
```

Constructing a matrix column-wise:

```{r}
m2 = matrix(1:6,
            nrow = 2,
            ncol = 3)

print(m2)
```

Can be creating by dimensions after a vector:

```{r}
m = 1:10
dim(m) = c(2,5)
print(m)
```

Matrices can also be created by binding stuff:

```{r}
x = 1:3
y = 10:12
cbind(x, y)
```
```{r}
rbind(x,y)
```

### Data types - Factors

Special type of vector to represent categorical data, examples:

- Male and female
- Treatment and control

They are used by special functions such as `lm()`, `DESeq()`, and `glm()`. Their use is encouraged since factors are self-describing.

#### factor() - creating factors

Factors are created as such:

```{r}
x = factor(c("yes", "yes", "no", "yes", "no"))
print(x)
```
Can be shown as tables:
```{r}
table(x)
```
Classes can be stripped from vectors:
```{r}
unclass(x)
```

#### levels() - ordering factors

Factors can be ordered, which is important for linear modelling, especially determining the baseline. Let's set up yes for the baseline:

```{r}
x = factor(c("yes", "yes", "no", "yes", "no"),
           levels = c("yes", "no"))
print(x)
```

### Missing values

Missing valures are either `NA` or `NaN` for math. Some functions might help with checking, such as:

- `is.na()` tests for NAs
- `is.nan()` check for NaNs
- `NA` values also have classes (int, char, etc)
- `NaN` is also `NA` but not the converse

#### is.na() - tests for missing values

```{r}
x = c(1, 2, NA, 10, 3)
is.na(x)
```

```{r}
is.nan(x)
```
And again, with `NaN`:
```{r}
x = c(1, 2, NaN, NA, 4)
is.na(x)
```
```{r}
is.nan(x)
```

### Data types - data.frames

Store tabular data, but with many more functionalities than a simple matrix despite being always rectangular

- Special type of list (every element has the same length)
- Each element is a column, with a row as length
- Can store almost anything
- Data frames have `row.names()` and `colNames()`
- Most common form of importing data
- Can be coerced to a matrix using `data.matrix()`

#### data.frame() - building DFs

```{r}
x = data.frame(foo = 1:4, bar = c(T, T, F, F))
print(x)
```
```{r}
nrow(x)
```
```{r}
ncol(x)
```

### Data types - attribute: Names

Very useful for writing readable code:

#### names() - naming things

```{r}
x = 1:3
names(x)
```
Above, there are no names, which can be added as such:
```{r}
names(x) = c("foo", "bar", "norf")
names(x)
```
And then, looking at `x` again:
```{r}
print(x)
```

Lists can also have names:

```{r}
x = list(
  a = 1,
  b = 2,
  c = 3
)

print(x)
```

And also matrices, with `dimnames()`:

```{r}
# Creates the matrix
m = matrix(1:4,
           nrow = 2,
           ncol = 2)

# Give names to rows and columns
dimnames(m) = list(c("a", "b"), c("c", "d"))
print(m)
```

## Reading tabular data

Primary functions to acquire data:

- `read.table()` and `read.csv()` are most common
- `readLines()` reads text files line by line
- `source()` reads R code
- `dgt()` reads deparsed R code
- `load()` recovers saved workspaces
- `unserialize()` reads single R objects in binary

### Writing data

The analogous output functions are:

- `write.table()`
- `writeLines()`
- `dump()`
- `dput()`
- `save()`
- `serialize()`

#### read.table() - most common data reader

Important paramaters are:

- `file` - the path of the file
- `header` - has header or not
- `sep` - how columns are separated
- `colClasses` - a char vector indicating the class of each column
- `nrows` - the number of rows
- `comment.char` - a string indicating a comment char
- `skip` - whether to skip any lines from the beginning
- `stringsAsFactors` - should variables be coded as factors?

### Reading large tables

With large sets, the lecturer gives some tips (think about RAM, remove default comment.char), etc. `colClasses` will be very important (2X speed) because the method of acquiring the input will not need to be tested or coerced into a larger generalized type. One such way of acquiring the classes is:


```{r eval=FALSE, include=TRUE}
pre = read.table("example.txt", nrows = 100)
classes = sapply(initial, class)
table_input = read.table("example.txt", colClasses = classes)
```

Also, keep in mind sysinfo:

- Available memory
- Other apps running
- Other users
- OS and architecture

### Textual data formats for storing data

Advocates for storing data in these formats (which are awesome btw)

- They're editable!
- `dump` and `dput` functions results in an editable text file while also preserving metadata
- Textual formats are great with version control
- Textual formats adhere to Unix philosophy
- Downside: Not space efficient

#### dputting R-objects

Pass data around on R by deparsing and reading back.

```{r}
y = data.frame(a = 1, b = "a")
dput(y)
```

Saving to a file with the object structure and all:

```{r}
dput(y, file = "./y.R")
new.y = dget("y.R")
new.y
```

I wonder if this can be done for some more complex stuff that can be read by other R in other places.


### Dumping R objects

Just as dput, but dump can put out namy objects in a single file:

```{r}
# Two random objects
x = "foo"
y = data.frame(a = 1, b = "a")

# Create dumpfile
dump(c("x", "y"), file = "data.R")

# Remove objects from memory
rm(x, y)

# Reads back as source file
source("data.R")
print(c(x,y))
```

### Connections

Data can be read in connection interfaces, with common or exotic file formats such as:

- file;
- compressed files such as gzfile or bzfile;
- url (Web connection)

The file function is:

```{r}
str(file)
```

The `open` part corresponds to code to do things such as:

- "r": read only
- "w": writing
- "a": appending
- "rb", "wb", "ab" - reading, writing or appending binary (Windows OS)

Connections let you navigate files or external objects, usually not needed directly. 

#### Reading webpages

Also with connections:

```{r}
con = url("http://www.jhsph.edu", "r")
x = readLines(con)
head(x, 12)
```

## Subsetting

Some operators:

- "[" always returns an object of the same class;
- "[[" extracts elements of a list or DF, only for a single element
- "$" extracts elements by name

Some primary examples:

```{r}
x = c("a", "b", "c", "c", "d", "a")
x[1]
```
```{r}
x[2]
```
```{r}
x[1:4]
```

Subsets with logical assignments:

```{r}
x[x > "a"]
```

Logical elements may also be assigned by testing the vector directly:

```{r}
u = x > "a"
u
```

## Subsetting lists

Double bracket and dollar sign operators are in rule.

```{r}
x = list(foo = 1:4, bar = 0.6)
x[1]
```
```{r}
x[[1]]
```
```{r}
x$bar
```
```{r}
x[["bar"]]
```
```{r}
x["bar"]
```

Multiple elements are extracted with the single bracket operator:

```{r}
x = list(
  foo = 1:4,
  bar = 0.6,
  baz = "hello"
)

x[c(1,3)]
```

The double bracket can compute indices:

```{r}
name = "foo"
x[[name]]
```
With single brackets:
```{r}
x$name
```
Again, with names:
```{r}
x$foo
```

Nested elements may also be extracted with "[[":

```{r}
x = list(
  a = list(10, 12, 14),
  b = c(3.14, 2.81)
)

# First index, 3rd element
x[[c(1,3)]]
```
```{r}
x[[1]][[3]]
```
```{r}
x[[c(2, 1)]]
```


## Subsetting matrices

Can be subsetted using `m(i,j)` linear algebra notation:

```{r}
x = matrix(1:6, 2, 3)
x[1, 2]
```
```{r}
x[2, 1]
```

Indices can also be missing, and will be understood as "all":

```{r}
x[1, ]
```
```{r}
x[, 2]
```

When a single element is retrieved, it returns as a vector of 1 rather then a 1x1 matrix, which can be changed by:

```{r}
x = matrix(1:6, 2, 3)
x[1, 2]
```
```{r}
x[1, 2, drop = F]
```

Also, when subsetting a single row or column, a vector is returned:

```{r}
x = matrix(1:6, 2, 3)
x[1, ]
```
Which can also be overrun by `drop = F`:
```{r}
x[1, , drop = 0]
```

## Partial matching

A tool that is allowed with "[[" and "$"

```{r}
x = list(aardvark = 1:5)
x$a
```
```{r}
x[["a"]]
```
```{r}
x[["a", exact = FALSE]]
```

## Removing NA values

A common task on data science.

```{r}
x = c(1, 2, NA, 4, NA, 5)
bad = is.na(x)
notbad = x[!bad]
print(notbad)
```

Subset of complete cases for 2+ objects is also possible with `complete.cases()`:

```{r}
x = c(1, 2, NA, 4, NA, 5)
y = c("A", "B", NA, "D", NA, "F")
cc = complete.cases(x, y)
cc
```
```{r}
x[cc]
```
```{r}
y[cc]
```

Again, with "airquality" data, subsetting two times::

```{r}
airquality[1:6, ]
```
```{r}
cc = complete.cases(airquality)
airquality[cc, ][1:6, ]
```

## Vectorized operations

- Removes many looping
- Feature shared with MATLAB
- Things to happen in parallel

Example of vectorized operation:

```{r}
x = 1:4
y = 6:9
x + y
```
```{r}
x > 2
```
```{r}
x >= 2
```
```{r}
y == 8
```
```{r}
x / y
```

# Quiz and Swirl exercises





