---
title: 'Week01: Overview'
output:
  html_document: 
    toc: yes
  pdf_document: default
  word_document: default
---

## Preamble

### Firing up python for Rmd

```{r}
library(reticulate)
use_python("C:/Users/PICHAU/anaconda3/python.exe")
```

### Additional packages and notes

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

## History of R

- First written in Fortran, then in C
- Environment for basic analysis, and then programming

## Features i didn't know of

- Identical to S
- Run on anything (even Playstation 3)
- Constant releases
- User turns into a programmer

## Free software?

1. Run the program
2. Study the program and source code
3. Freedom to redistribute
4. Freedom to improve the program

## Drawbacks

1. 40-year old technology
2. 3D/dynamic graphics sucks
3. Functionality is based on demand
4. Objects are (usually) stored in physical memory
5. Slow if you don't know C++can 

## Asking questions

Pretty basic, i'm a master forumer already.

## Code: Input and Evaluation

```{r}
## Assignment operators
x = 1
y <- 2
print(x)
print(y)

## Incomplete expression
# ex:
# x <-
```

Evaluation: whenever some object is called, it is auto-printed:

```{r}
print(x)
```

Let's try a sequence:

```{r}
seq = 1:20
print(seq)
```


## R Objects and Attributes

- Everything is an object
- Atomic types:
  - Char
  - Num
  - Int
  - Complex
  - Logicals
- The most basic object is a vector
- Lists are represented as vector, but may contain anything inside

### Numbers

- Num objects
- An integer must be specified with L, such as: `1L`
- Other numbers:
  - `Inf` is infinity
  - `NaN` is "not a number", or missing value
  
### Attributes

R objects can have attributes:

- names, dimnames
- dimensions
- class
- length
- anything user-defined
- check with `attributes()`

### Data types: vectors and functions

#### c() - concatenates values

Creates vectors with the desired content

```{r}
x = c(0.5, 0.6)
x = c(2L, 4L)
x = 9:29
x = c(1+0i, 2+4i)
```

#### vector() - creates vectors

```{r}
x = vector("numeric", length = 15)
print(x)
```

#### Mixing objects and coercion

when mixing objects in a vector, the least enforced type is coerced, from boolean, to integer, to character

#### Explicit coercion: as.type()

We can force a class to be whatever type we want explicitly:

```{r}
x = 0:5
as.numeric(x)
as.logical(x)
as.character(x)
as.complex(x)
```

Nonsensical coercion results in NAs:

```{r}
x = c("a","b","c")
as.numeric(x)
as.logical(x)
as.complex(x)
```

### Lists

Super vectors that can carry anything as an indexed element

```{r}
x = list(1, "a", T, 1+4i)
print(x)
```

### Matrices

Vectors with a dimension attribute, an integer

```{r}
m = matrix(nrow = 2, ncol = 3)
m
```
```{r}
dim(m)
```
```{r}
attributes(m)
```

Constructing a matrix column-wise:

```{r}
m2 = matrix(1:6,
            nrow = 2,
            ncol = 3)

print(m2)
```

Can be creating by dimensions after a vector:

```{r}
m = 1:10
dim(m) = c(2,5)
print(m)
```

Matrices can also be created by binding stuff:

```{r}
x = 1:3
y = 10:12
cbind(x, y)
```
```{r}
rbind(x,y)
```

### Data types - Factors

Special type of vector to represent categorical data, examples:

- Male and female
- Treatment and control

They are used by special functions such as `lm()`, `DESeq()`, and `glm()`. Their use is encouraged since factors are self-describing.

#### factor() - creating factors

Factors are created as such:

```{r}
x = factor(c("yes", "yes", "no", "yes", "no"))
print(x)
```
Can be shown as tables:
```{r}
table(x)
```
Classes can be stripped from vectors:
```{r}
unclass(x)
```

#### levels() - ordering factors

Factors can be ordered, which is important for linear modelling, especially determining the baseline. Let's set up yes for the baseline:

```{r}
x = factor(c("yes", "yes", "no", "yes", "no"),
           levels = c("yes", "no"))
print(x)
```

### Missing values

Missing valures are either `NA` or `NaN` for math. Some functions might help with checking, such as:

- `is.na()` tests for NAs
- `is.nan()` check for NaNs
- `NA` values also have classes (int, char, etc)
- `NaN` is also `NA` but not the converse

#### is.na() - tests for missing values

```{r}
x = c(1, 2, NA, 10, 3)
is.na(x)
```

```{r}
is.nan(x)
```
And again, with `NaN`:
```{r}
x = c(1, 2, NaN, NA, 4)
is.na(x)
```
```{r}
is.nan(x)
```

### Data types - data.frames

Store tabular data, but with many more functionalities than a simple matrix despite being always rectangular

- Special type of list (every element has the same length)
- Each element is a column, with a row as length
- Can store almost anything
- Data frames have `row.names()` and `colNames()`
- Most common form of importing data
- Can be coerced to a matrix using `data.matrix()`


















