---
title: "Week 2 - Control Structures"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

Control structures!
Some testing with unusual forms:

### "If/else"

```{r}
## Some wild forms!
x <- 2
y <- if (x > 3) {10} else {0}
```

### "For" loops

```{r}
## Crunched version
for (i in 1:10) {print(i)}

## More spaced out
for (n in 1:15) {
  print(n)
}

```

C-like form:

```{r}
x <- c("a", "b", "c", "d")

for(i in 1:4) {print(x[i])}
```

```{r}
# With seq_along
for (i in seq_along(x)) { print(x[i]) }
```

Compact-style:

```{r}
for (letter in x) { print(letter) }
```

### Nested for loops

Bit of a pain to read, try to get around

```{r}
x <- matrix(1:6, 2, 3)

for (i in seq_len(nrow(x))) {
  for(j in seq_len(ncol(x))) {
    print(x[i, j])
  }
}
```

### While loops

Runs a condition while something is true, when false it breaks

- Stop condition (or infinites)

```{r}
count <- 0

while (count < 10) {
  print(count)
  count = count +1
}
```

Testing multiple conditions:

```{r}
z = 5

# Coin flip tests
while ( z >= 3 && z <= 10) {
  print(z)
  coin <- rbinom(1, 1, 0.5)
  
  if ( coin == 1 ) { z = z + 1 } 
  else { z = z - 1 }
    
}

```

### Repeat, next, break

- Infinite loop
- Not common in statistics but have uses
- Vectorizing is always usually better

BTW very badly written part, code doesn't work!

```{r eval=FALSE}
x0 = 1
tol = 1e-8

repeat {
  x1 <- computeEstimate()
  
  if (abs(x1-x0) < tol) {
    break
  } else {
    x0 = x1
  }
}

```

### Next & Return

- Next skips an interation
- Return signals an ending

```{r}
for (i in 1:100) {
  if (i <= 96) {
    next
  }
  print("blah")
}
```

## First function

- Suggestion: separate file

Sum two files:

```{r}
add2 <- function(x, y) {x + y}
```


### Function: Above10

- Returns numbers above 10

```{r}
above10 <- function(x) {
  use <- x > 10
  x[use]
}
```

### AboveN: Returns numbers above a given number

```{r}
aboveN <- function(x, n) {
  use <- x > n
  x[use]
}
```


### AboveN with a default argument

```{r}
aboveN <- function(x, n = 10) {
  use <- x > n
  x[use]
}
```

### ColMean

- removeNA doesn't work in this case

```{r}
colmean <- function(y, removeNA = T) {
  nc = ncol(y)
  means = numeric(nc)
  for (i in 1:nc) {
    means[i] <- mean(y[, i])
  }
  means
}
```

## Functions part I actualy II

- How functions speed things up
- Functions are first class object (aka can be used like anything else)
- Functions have named arguments, aka x in `function(x)`
- `formals()` returns all arguments of a function
- Arguments can be matched positionally or by name and can be also mixed (which is kinda useless and messy and i dont believe i've ever used it, or will)
- Functions can be partially matched!

## Functions part II actually III

- Default argument value can be `NULL`
- Functions have lazy evaluation: are only evaluated as they are needed as long as you pass NULL

```{r}
lazysqr = function(a = NULL, b = NULL) {
  a^2
  b^2
}
```

A printing funcion:

```{r}
printer = function(a, b = NULL) {
  print(a)
  print(b)
}
```

## The "..." Argument

- `...` indicates a variable number of args that are passed to other functions
- They extend another function in a situation in which you dont want to copy the entire argument list of the original function
- `...` can be also used when the number of arguments cannot be known in advance (thii about `paste()`, for instance)
- Any arguments after `...` must be named explicitly, and cannot be partially matched which is logical

Example: plots with lines (type instead of dot):

```{r}
myplot = function(x, y, type = "l", ...) {
  plot(x, y, type = type, ...)
}
```

## Symbol binding

- Named scopes and precedence
- Attribution happens by searching environments for appropriate values, in the order:

1. Anything entered by user (which enters the global envo, but takes precedence)
2. Search the global environment
3. Search the namespaces of the packages on the search list `search()`

- One can define which packages are loaded, in what order
- A package loaded with `library()` takes precedence over global.envo
- Namespaces are different for functions and non-functions
- R uses lexical/statical scoping, supposedly useful for statistics
- The scoping rules of a language determine how values are assigned to free valuables. In R, the values of free valuables are searched for in the environment that defines the function.
- A function (L1) inside a function (L2) has the environment of L1!

```{r}
## Function returns another function as output!
make.power <- function(n) {
  pow <- function(x) {
    x^n
  }
  pow
}

cube <- make.power(3)
sqrd <- make.power(2)
```

### Exploring a function closure

What's in a function environment?

```{r}
## Finds what's an environment
ls(environment(cube))
```

```{r}
## Gets detailed values
get("n", environment(cube))
```

### Lexical vs dynamic scoping

They are the same if you define everything on the global envo

```{r}
g = function(x) {
  a = 3
  x + a + y
}

g(2)
```

## Optimization

Bit of critique here: the content doesn't match the level of classes at all. Bit of a mismatch here.

- optim. routines requires as an arg a function whose argument is a vector of parameters
- An object function, however, might depend on other things besides its own parameters (such as stuff from the environment)
- When writing optimization software, sometimes its good to let the user fix certain parameters

## Dates and times

### Dates

```{r}
x = as.Date("1970-01-01")
x
```

```{r}
unclass(x)
```
```{r}
unclass(as.Date("1970-02-01"))
```

### Times

- POSIXct - very large integer
- POSIXlt - list that store a bunch of information

POSIXct is a pure integer. POSIXlt is readable:

```{r}
x = Sys.time()
p = as.POSIXlt(x)
names(unclass(p))
```

### strptime function captures a date in a different format

Code didnt work

```{r}
datestring = c("January 10, 2012 10:40", "December 9, 2011 9:10")
x = strptime(datestring, "%B %d, %Y %H:%M")
x
class(x)
```

### Operation on dates and times

Can add or subtract dates:

```{r}
x = as.Date("2012-01-01")
y = strptime("9 Jan 2011 11:34:21", "%d %b %Y %H:%M:%S")
x = as.POSIXlt(x)
x-y
```

Also difference in time zones!

```{r}
x = as.POSIXct("2012-10-25 01:00:00")
y = as.POSIXct("2012-10-25 06:00:00", tz = "GMT")
y-x
```

# Week 2 Quiz

1. 

```{r}
cube <- function(x, n) {
        x^3
}

cube(3)
```

2. The following code will produce a warning in R. (It did not!)

```{r}
x <- 1:10
if(x > 5) {x <- 0}
```

3. Consider the following function

```{r}
f <- function(x) {
        g <- function(y) {y + z}
        z <- 4
        x + g(x)
}

z = -15
f(3)
```

Consider the following expression. What is the value of 'y' after evaluating 
this expression?

```{r}

x <- 5
y <- if(x < 3) {NA} else {10}
```

# Programming assignment

- Passing working directory

```{r}
specdata <- "./02_RProg/Week02_data/specdata/"
pollutantmean(specdata, "sulfate", 1:10)

cr <- corr(specdata, 129)                
cr <- sort(cr)                
n <- length(cr)    
RNGversion("3.5.1")
set.seed(197)                
out <- c(n, round(cr[sample(n, 5)], 4))
print(out)

cr <- corr(specdata, 2000)                
n <- length(cr)                
cr <- corr(specdata, 1000)                
cr <- sort(cr)
print(c(n, round(cr, 4)))
```



