---
title: ""
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Week 4: Simulation and Profiling

## `str` displays the internal structure of an R object

- diagnostic alternative to summary
- well suited for compact display
- roughly one line per basic object

```{r}
str(str)
```

```{r}
str(lm)
```

```{r}
str(ls)
```

```{r}
x <- rnorm(100, 2, 4)
summary(x)
```
```{r}
str(x)
```

```{r}
f <- gl(40, 10)
str(f)
```

```{r}
library(datasets)
str(airquality)
```

```{r}
m <- matrix(rnorm(100), 10, 10)
str(m)
```

Splits a df and see its structure:

```{r}
s <- split(airquality, airquality$Month)
str(s)
```

## Simulation/RNG

Functions are:

- rnorm: random normal variates with mean and sd
- dnorm: eval the norm prob dens
- pnorm: eval the cumul distr
- rpois: generate rand poisson variates with a rate

For any given distributions there are 04 functions:

- d for density
- r for RNG
- p for cumulative
- q for quantile

### Seeds

Can be set to fix values for random numbers:

```{r}
set.seed(1)
rnorm(5)
```

### Generation Poisson data

```{r}
rpois(10, 1)
```
```{r}
rpois(10, 2)
```

```{r}
rpois(10, 20)
```


```{r}
## Cumulative
ppois(2, 2)
```

```{r}
ppois(4, 2)
```

```{r}
ppois(6, 2)
```

## Simulating a linear model

- One predictor (x)
- Noise (epsilon, sd = 2)

```{r}
set.seed(20)
x <- rnorm(100)
e <- rnorm(100, 0, 2)
y <- 0.5 + 2*x + e
summary(y)
```

```{r}
plot(x, y)
```

### Random model with a binary X

```{r}
set.seed(10)
x <- rbinom(100, 1, 0.5)
e <- rnorm(100, 0, 2)
y <- 0.5 + 2*x + e
summary(y)
```
```{r}
plot(x, y)
```

### RNG form a GLM

```{r}
set.seed(1)
x <- rnorm(100)
log.mu <- 0.5 + 0.3 * x
y <- rpois(100, exp(log.mu))
summary(y)
```
```{r}
plot(x, y)
```

## Random sampling

- Sample draws randomly from a vector

```{r}
set.seed(1)
sample(1:10, 4)
```
```{r}
sample(1:10, 4)
```
```{r}
sample(letters, 5)
```
```{r}
sample(1:10)
```
```{r}
sample(1:10, replace = T)
```

## R Profiler

- Large programs
- Figures why it takes so much time
- Find large, dumb loops
- Guess first
- Profile then
- Dont optimize prematurely, write correct code, then optimize

### system.time()

- Evaluates an expression and times it
- Returns an object of class `proc_time`
  - user time: time charged to CPU
  - elapsed time: "wall clock" time
- UT > ET - parallel processing
- UT < ET - CPU in idle/waiting

Some examples
```{r}
## Reading pages are slow
system.time(readLines("http://www.jhsph.edu"))
```

```{r}
hilbert <- function(n) {
  i <- 1:n
  1 / outer(i - 1, i, "+")
}

## Faster for MAC, it is!
x <- hilbert(1000)
system.time(svd(x))
```

- Even longer expressions by just wrapping them in sys.time.

## R profiler: Rprof()

- R must be compiled with prof.support
- `summaryRprof()` summarizes the output
- Do not use with systime!
- Tracks the call stack in intervals and tabulates the time
- Default timing: 0.02s
- Useless for very fast functions
- by.total: 100% on the top level (sometimes useless), the helper that do the real work is less than 100%.
- Subtract the work of the lower level function is very interesting


# Week 4 quiz

1. What is produced at the end of this snippet of R code?

```{r}
set.seed(1)
rpois(5, 2)
```


5. What does it do?

```{r}
set.seed(10)
x <- rep(0:1, each = 5)
e <- rnorm(10, 0, 20)
y <- 0.5 + 2 * x + e
```









